A Fundamental Measure Theory Functional for Hard-Sphere Contact Densities

### 2012-07-17 papers/contact/paper.tex
Added correlation in SomeFunctionals.hs small changes to Expresions.lhs, fixed a few things in Reciprical Shell prime functions Tested everything on node02.

\begin{align}
  p_{HS}(\mathbf{r}) &= \frac{1}{n_0(\rr) 4\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta R(\mathbf{r})} \label{eq:p_{HS}^S}\\
  \ncontact(\rr) &= \frac{1}{n_0(\rr) k_BT 4\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta R(\mathbf{r})} \\
  g_\sigma^A(\rr) &= \frac{\ncontact(\rr)}{n_0(\rr)} \\
  &= \frac{1}{n_0(\rr)^2}\frac{1}{ k_BT 4\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta R(\mathbf{r})} \label{eq:g-A-exact}
\end{align}
where the functional derivative $\frac{\delta A_{HS}}{\delta
  R(\mathbf{r})}$ in this case is computed with the symmetric
formulation of the fundamental measures.  Complete expressions for
$g_\sigma^S$ are included in Appendix~\ref{appendix:g-S}.\\

@@

\newcommand\sphereExplanation{ Blue curves describe the density of
  hard spheres.  Green curves describe the contact density averaged
  according to the point of contact.  Red curves describe the contact
  density as averaged according to the centers of the spheres that are
  touching.  In each case, solid lines correspond to the Monte Carlo
  simulation, and dashed lines of various sorts represent DFT model
  predictions.  }

@@

The derivatation of the asymmetrically averaged correlation function
$g_\sigma^A(\rr)$ requires that we express the free energy in terms of
a set of weighted densities in which the radius of the weighting
kernel is determined according to the position of the sphere in
question---which is to say, the location where the density itself is
evaluated.  This is the most natural way to construct such a
functional, in which the radius of spheres is a function of position.
Thus we define $n_3^A$ as
\begin{align}
  n_3^{A}(\rr) &= \int n(\rr') \Theta(\left|\rr - \rr'\right| -R(\rr')) d\rr'
\end{align}
and we define the other weighted densities similarly.  In order to
compute derivatives of the free energy with respect to $R(\rr)$, we
need to find $g_\sigma^A$.  In the case of $n_3^A$, we needS
\begin{align}
  \frac{\delta n_3^{A} (\rr')}{\delta R(\rr)} &=
  \int n (\rr'') \delta(|\rr' - \rr''| - R(\rr'')) \delta(\rr-\rr'') d\rr'' \\
   &= n (\rr) \delta(|\rr' - \rr| - R(\rr))
\end{align}
Similarly, for $n_2^A$, we find
\begin{align}
  n_2^{A}(\rr') &= \int n(\rr'') \delta(|\rr' - \rr''| - R(\rr'')) d \rr''\\
  \frac{\delta n_2^{A}(\rr')}{\delta R(\rr)} &= -n(\rr) \delta'(|\rr' - \rr| - R(\rr))\\
\end{align}
This is somewhat more challenging, since it involves an unfamiliar
function, the derivative the Dirac delta function $\delta'$.
%
The derivatives of the remaining scalar densities can be reduced to
sums of the terms above:
\begin{align}
  n_1(\rr') &= \int \mathbf{dr''} \frac{n(\rr'')}{4\pi R(\rr'')}
  \delta(|\rr'-\rr''| - R(\rr'')) \\
  \frac{\delta n_1(\rr')}{\delta R(\rr)}
  &= -\frac{n(\rr)}{4\pi
    R(\rr)}\delta'(|\rr'-\rr| - R(\rr))
  -
  \frac{n(\rr)}{4\pi
    R(\rr)^2}\delta(|\rr'-\rr| - R(\rr))
\end{align}
and
\begin{align}
  n_0(\rr') &= \int \mathbf{dr''} \frac{n(\rr'')}{4\pi R(\rr'')^2}
  \delta(|\rr'-\rr''| - R(\rr'')) \\
  \frac{\delta n_1(\rr')}{\delta R(\rr)^2}
  &= -\frac{n(\rr)}{4\pi
    R(\rr)^2}\delta'(|\rr'-\rr| - R(\rr))
  -
  2\frac{n(\rr)}{4\pi
    R(\rr)^3}\delta(|\rr'-\rr| - R(\rr))
\end{align}
The vector-weighted densities have no additional occurrences of $R$,
their functional derivatives come out similarly:
\begin{align}
  \mathbf{n}_{V2}^{A}(\rr) &= \int n(\rr') \delta(|\rr' - \rr''| - R(\rr''))
    \frac{\rr'-\rr''}{|\rr'-\rr''|} d \rr''\\
  \frac{\delta \mathbf{n}_{V2}^{A}(\rr')}{\delta R(\rr)} &= - n(\rr) \delta'(|\rr' - \rr| - R(\rr))
    \frac{\rr'-\rr}{|\rr'-\rr|}\\
\end{align}
and analogously for $\mathbf{n}_{V1}^A$:
\begin{align}
  \mathbf{n}_{V1}^A(\rr) &= \int d\rr' \frac{n(\rr')}{4\pi R(\rr')}
  \delta(|\mathbf{r}-\rr'| - R(\rr')) \frac{\rr-\rr'}{|\rr-\rr'|}\\
  \frac{\delta \mathbf{n}_{V1}^A(\rr')}{\delta R(\rr)}
  &= -\frac{n(\rr)}{4\pi
    R(\rr)}\delta'(|\rr'-\rr| - R(\rr)) \frac{\rr'-\rr}{|\rr'-\rr|}
  -
  \frac{n(\rr)}{4\pi
    R(\rr)^2}\delta(|\rr'-\rr| - R(\rr)) \frac{\rr'-\rr}{|\rr'-\rr|}
\end{align}
\fixme{double-check the directions of all the vector quantities...}

@@

The functional derivative of $n_2^S(\rr')$ is quite similar:
\begin{align}
  n_2^{S}(\rr) &= \int n(\rr') \delta(|\rr - \rr'| + R(\rr))d\mathbf
  r'\\
  \frac{\delta n_2^{S}(\rr')}{\delta R(\rr)} &= -\delta(\rr-\rr') \int n(\rr'')
  \delta'(|\rr'-\rr''| - R(\rr')) d\rr''
\end{align}
where we recognize that the convolution---while not one of the standard
FMT weighted density---is a simple convolution of the density.  We can
go on to compute the derivatives of the remaining weighted densities:
\begin{align}
  n_1^{S}(\rr) &= \int \frac{n(\rr')}{4\pi R(\rr)} \delta(|\rr - \rr'| + R(\rr))d\mathbf r'\\
    \frac{\delta n_1^{S}(\rr')}{\delta R(\rr)} &=
    -\delta(\rr-\rr') \left( \int \frac{n(\rr'')}{4\pi R(\rr)}
    \delta'(|\rr'-\rr''| - R(\rr')) d\rr''
    + n_0^s(\rr') \right)
\end{align}
\begin{align}
  n_0^{S}(\rr) &= \int \frac{n(\rr')}{4\pi R(\rr)^2} \delta(|\rr - \rr'| + R(\rr))d\mathbf r'\\
    \frac{\delta n_0^{S}(\rr')}{\delta R(\rr)}
    &= -\delta(\rr-\rr') \left( \int \frac{n(\rr'')}{4\pi R(\rr)^2}
    \delta'(|\rr'-\rr''| - R(\rr')) d\rr''
    +
    \frac{2}{R} n_0^S(\rr') \right)
\end{align}
\begin{align}
  \mathbf{n}_{V2}^{S}(\rr) &= \int n(\rr') \delta(|\rr - \rr'| - R(\rr))
  \frac{(\rr - \rr')}{|\rr - \rr'|} d \rr'\\
  \frac{\delta \mathbf{n}_{V2}^{S}(\rr')}{\delta R(\rr)} &= \delta(\rr - \rr')
  \int n(\rr'') \delta'(|\rr' - \rr''| - R(\rr'))
  \frac{(\rr - \rr'')}{|\rr - \rr''|} d\rr''
\end{align}
\begin{align}
 \mathbf{n}_{V1}^{S}(\rr) &= \int \frac{n(\rr')}{4\pi R(\rr)} \delta(|\rr - \rr'| + R(\rr))d\mathbf r'
   \frac{(\rr - \rr')}{|\rr - \rr'|}\\
 \frac{\delta \mathbf{n}_{V1}^{S}(\rr')}{\delta R(\rr)}
 &= -\delta(\rr-\rr')\left(\int \frac{n(\rr'')}{4\pi R(\rr)}
   \delta'(|\rr'-\rr''| - R(\rr')) \frac{(\rr - \rr'')}{|\rr - \rr''|}  d\rr''
   + \mathbf{n}_{V2}^S(\rr') \right)
\end{align}

>>>>

\begin{align}
  p_{HS}(\mathbf{r}) &= \frac{1}{n_0(\rr) 4\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta R(\mathbf{r})} \label{eq:p_{HS}^S}\\
  \ncontact(\rr) &= \frac{1}{n_0(\rr) k_BT 4\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta R(\mathbf{r})} \\
  g_\sigma^S(\rr) &= \frac{\ncontact(\rr)}{n_0(\rr)} \\
  &= \frac{1}{n_0(\rr)^2}\frac{1}{ k_BT 4\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta R(\mathbf{r})} \label{eq:g-A-exact}
\end{align}
where the functional derivative $\frac{\delta A_{HS}}{\delta
  R(\mathbf{r})}$ in this case is computed with the symmetric
formulation of the fundamental measures.  Complete expressions for
$g_\sigma^S$ are included in Appendix~\ref{appendix:g-S}.\\

@@

\newcommand\sphereExplanation{ Blue curves describe the density of
  hard spheres.  Green curves describe the contact density averaged
  according to the point of contact.  Red curves describe the contact
  density as averaged according to the centers of the spheres that are
  touching.  In each case, solid lines correspond to the Monte Carlo
  simulation, and dashed lines of various sorts represent DFT model
  predictions.}

@@

The derivatation of the asymmetrically averaged correlation function
$g_\sigma^A(\rr)$ requires that we express the free energy in terms of
a set of weighted densities in which the radius of the weighting
kernel is determined according to the position of the sphere in
question---which is to say, the location where the density itself is
evaluated.  This is the most natural way to construct such a
functional, in which the radius of spheres is a function of position.
Thus we define $n_3^A$ as
\begin{align}
  n_3^{A}(\rr) &= \int n(\rr') \Theta(\left|\rr - \rr'\right| -R(\rr')) d\rr'
\end{align}
and we define the other weighted densities similarly.  In order to
compute derivatives of the free energy with respect to $R(\rr)$, we
need to find $g_\sigma^A$.  In the case of $n_3^A$, we needS
\begin{align}
  \frac{\delta n_3^{A} (\rr')}{\delta R(\rr)} &=
  \int n (\rr'') \delta(|\rr' - \rr''| - R(\rr'')) \delta(\rr-\rr'') d\rr'' \\
   &= n (\rr) \delta(|\rr' - \rr| - R(\rr))
\end{align}
Similarly, for $n_2^A$, we find
\begin{align}
  n_2^{A}(\rr') &= \int n(\rr'') \delta(|\rr' - \rr''| - R(\rr'')) d \rr''\\
  \frac{\delta n_2^{A}(\rr')}{\delta R(\rr)} &= -n(\rr) \delta'(|\rr' - \rr| - R(\rr))\\
\end{align}
This is somewhat more challenging, since it involves an unfamiliar
function, the derivative the Dirac delta function $\delta'$.
%
The derivatives of the remaining scalar densities can be reduced to
sums of the terms above:
\begin{align}
  n_1(\rr') &= \int \mathbf{dr''} \frac{n(\rr'')}{4\pi R(\rr'')}
  \delta(|\rr'-\rr''| - R(\rr'')) \\
  \frac{\delta n_1(\rr')}{\delta R(\rr)}
  &= -\frac{n(\rr)}{4\pi
    R(\rr)}\delta'(|\rr'-\rr| - R(\rr))
  -
  \frac{n(\rr)}{4\pi
    R(\rr)^2}\delta(|\rr'-\rr| - R(\rr))
\end{align}
and
\begin{align}
  n_0(\rr') &= \int \mathbf{dr''} \frac{n(\rr'')}{4\pi R(\rr'')^2}
  \delta(|\rr'-\rr''| - R(\rr'')) \\
  \frac{\delta n_0(\rr')}{\delta R(\rr)}
  &= -\frac{n(\rr)}{4\pi
    R(\rr)^2}\delta'(|\rr'-\rr| - R(\rr))
  -
  2\frac{n(\rr)}{4\pi
    R(\rr)^3}\delta(|\rr'-\rr| - R(\rr))
\end{align}
The vector-weighted densities have no additional occurrences of $R$,
their functional derivatives come out similarly:
\begin{align}
  \mathbf{n}_{V2}^{A}(\rr) &= \int n(\rr') \delta(|\rr' - \rr''| - R(\rr''))
    \frac{\rr'-\rr''}{|\rr'-\rr''|} d \rr''\\
  \frac{\delta \mathbf{n}_{V2}^{A}(\rr')}{\delta R(\rr)} &= - n(\rr) \delta'(|\rr' - \rr| - R(\rr))
    \frac{\rr'-\rr}{|\rr'-\rr|}\\
\end{align}
and analogously for $\mathbf{n}_{V1}^A$:
\begin{align}
  \mathbf{n}_{V1}^A(\rr) &= \int d\rr' \frac{n(\rr')}{4\pi R(\rr')}
  \delta(|\mathbf{r}-\rr'| - R(\rr')) \frac{\rr-\rr'}{|\rr-\rr'|}\\
  \frac{\delta \mathbf{n}_{V1}^A(\rr')}{\delta R(\rr)}
  &= -\frac{n(\rr)}{4\pi
    R(\rr)}\delta'(|\rr'-\rr| - R(\rr)) \frac{\rr'-\rr}{|\rr'-\rr|}
  -
  \frac{n(\rr)}{4\pi
    R(\rr)^2}\delta(|\rr'-\rr| - R(\rr)) \frac{\rr'-\rr}{|\rr'-\rr|}
\end{align}
\fixme{double-check the directions of all the vector quantities...}

@@

The functional derivative of $n_2^S(\rr')$ is quite similar:
\begin{align}
  n_2^{S}(\rr) &= \int n(\rr') \delta(|\rr - \rr'| - R(\rr))d\mathbf
  r'\\
  \frac{\delta n_2^{S}(\rr')}{\delta R(\rr)} &= -\delta(\rr-\rr') \int n(\rr'')
  \delta'(|\rr'-\rr''| - R(\rr')) d\rr''
\end{align}
where we recognize that the convolution---while not one of the standard
FMT weighted density---is a simple convolution of the density.  We can
go on to compute the derivatives of the remaining weighted densities:
\begin{align}
  n_1^{S}(\rr) &= \int \frac{n(\rr')}{4\pi R(\rr)} \delta(|\rr - \rr'| - R(\rr))d\mathbf r'\\
    \frac{\delta n_1^{S}(\rr')}{\delta R(\rr)} &=
    -\delta(\rr-\rr') \left( \int \frac{n(\rr'')}{4\pi R(\rr)}
    \delta'(|\rr'-\rr''| - R(\rr')) d\rr''
    + n_0^s(\rr') \right)
\end{align}
\begin{align}
  n_0^{S}(\rr) &= \int \frac{n(\rr')}{4\pi R(\rr)^2} \delta(|\rr - \rr'| - R(\rr))d\mathbf r'\\
    \frac{\delta n_0^{S}(\rr')}{\delta R(\rr)}
    &= -\delta(\rr-\rr') \left( \int \frac{n(\rr'')}{4\pi R(\rr)^2}
    \delta'(|\rr'-\rr''| - R(\rr')) d\rr''
    +
    \frac{2}{R} n_0^S(\rr') \right)
\end{align}
\begin{align}
  \mathbf{n}_{V2}^{S}(\rr) &= \int n(\rr') \delta(|\rr - \rr'| - R(\rr))
  \frac{(\rr - \rr')}{|\rr - \rr'|} d \rr'\\
  \frac{\delta \mathbf{n}_{V2}^{S}(\rr')}{\delta R(\rr)} &= -\delta(\rr - \rr')
  \int n(\rr'') \delta'(|\rr' - \rr''| - R(\rr'))
  \frac{(\rr - \rr'')}{|\rr - \rr''|} d\rr''
\end{align}
\begin{align}
 \mathbf{n}_{V1}^{S}(\rr) &= \int \frac{n(\rr')}{4\pi R(\rr)} \delta(|\rr - \rr'| + R(\rr))d\mathbf r'
   \frac{(\rr - \rr')}{|\rr - \rr'|}\\
 \frac{\delta \mathbf{n}_{V1}^{S}(\rr')}{\delta R(\rr)}
 &= -\delta(\rr-\rr')\left(\int \frac{n(\rr'')}{4\pi R(\rr)}
   \delta'(|\rr'-\rr''| - R(\rr')) \frac{(\rr - \rr'')}{|\rr - \rr''|}  d\rr''
   + \frac{\mathbf{n}_{V2}^S(\rr')}{4\pi R(\rr)^2} \right)
\end{align}
